---
title: "Housing Affordability in Toronto"
subtitle: "Visualizing 2021 Census Data"
author: "Fraser McLaughlin"
contact: 'fraser.mclaughlin@outlook.com'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    toc: yes
    df_print: paged
    toc_float: yes
    number_sections: yes
    code_folding: hide
    
---
```{css, echo=FALSE}

p.caption {
  font-size: 0.8em;
  font-weight: bold;
}

.customTable {
  width: 100%;
  border: 1px solid black;
  border-collapse: collapse;
  
.customTable th,
.customTable td {
  width: 20%;
  text-align: left;
  padding: 4px;
  border: 1px solid black;
  border-collapse: collapse;
}
}
```

```{r setup, warning=FALSE, echo=FALSE}
library("pacman")
pacman::p_load(
  tidyverse,
  scales,
  leaflet,
  plotly,
  reactable,
  htmltools,
  htmlwidgets,
  leaflet.extras
)

```


# Introduction

This report provides information on the spatial distribution of housing affordability in 
Toronto based on 2021 Census data. This is a personal interest 
project and a work in progress.

The source code for data acquisition, visualization, and creating this report are 
all [available on GitHub](https://github.com/frasemcl/kensignton-market).

This began as a data pull for the 
[Kensington Market Community Land Trust](https://kmclt.ca/), an organization
I recommend checking out, and turned into a deeper dive into the results from the 
most recent Census from a Toronto housing affordability perspective.

When I find time to take this further, there are many more 2021 variables 
that I'd like to investigate, but my top priority is to demonstrate how the variables
shown below have changed over time.


# Data

- 2021 Census results and Census Tract (CT) boundaries are from the 
[cancensus R package](https://mountainmath.github.io/cancensus/).

- Toronto neighbourhood boundaries are from the [opendatatoronto R package](https://cran.r-project.org/web/packages/opendatatoronto/index.html)

Each CT name (which is actaully a number) has been joined to 
the Toronto neighbourhood that it is located in. The process of naming the CTs based
on the neighbourhood is outlined in [the appendix](#naming-cts).

For example, the CT that delineates Kensignton Market is called '0038.00' and in 
this analysis it has been renamed **'0038.00_Kensington-Chinatown'**. To evaluate 
Kensington-Chinatown neighbourhood as a whole, you would need to consider multiple CTs:
'0036.00_Kensington-Chinatown', '0037.00_Kensington-Chinatown', '0038.00_Kensington-Chinatown',
'0039.00_Kensington-Chinatown'

# Visualization
## General Info
This table contains some basic demographic information and income information for
all 585 Toronto Census Tracts. For more full details on each census variable,
consult [this lookup table in the appendix](#var-lookup).

### Table
**v_CA21_1:** Population, 2021\
**v_CA21_3:** Population percentage change, 2016 to 2021\
**v_CA21_386:** Average age\
**v_CA21_560:** Median total income in 2020 among recipients (\$)\
**v_CA21_906:** Median total income of household in 2020 (\$)\

**To sort by a variable, click on the variable name**\
**Search by CT or neighbourhood name for details by CT**
```{r, warning=FALSE,echo=FALSE}
table_background_info
```

## Expensive Housing
Households spending more than 30% of total houshold income on shelter costs,
be it rent, a mortgage, or repairs, are considered to be facing expensive 
housing costs. The following map and table provide interactive means to explore 
Census variables describing expensive housing across the city.

### Map
**v_CA21_4315:** % of tenant households spending 30% or more of its income on shelter costs\
**v_CA21_4307:** % of owner households spending 30% or more of its income on shelter costs\

**Hover, click CT for more info, zoom in/out, and toggle between the two choropleth layers (top right)**
```{r, echo=FALSE}
map_leaflet_perc30
```

### Table
**v_CA21_4290:** (households) Spending 30% or more of income on shelter costs\
**v_CA21_4307:** % of owner households spending 30% or more of its income on shelter costs\
**v_CA21_4315:** % of tenant households spending 30% or more of its income on shelter costs\

**To sort by a variable, click on the variable name**
```{r, warning=FALSE,echo=FALSE}
table_expensive
```

## Housing Density
The following map and table can be used to explore the amount of housing in each
Census Tract.

### Map
**v_CA21_434:** Total Occupied dwellings\
**v_CA21_452:** Total Average household size\

**Hover, click CT for more info, zoom in/out, and toggle between the two choropleth layers (top right)**
```{r, echo=FALSE}
map_leaflet_house_amnt
```

### Table
For more full details on each census variable consult [the lookup table in the appendix](#var-lookup).

**v_CA21_434:** Total Occupied dwellings\
**v_CA21_452:** Total Average household size\

**To sort by a variable, click on the variable name**
```{r, warning=FALSE,echo=FALSE}
table_housing_amt
```

## Housing Cost
The following map and table contain data on housing cost by CT.

### Map

**v_CA21_4309:** Median monthly shelter costs for owned dwellings (\$)\
**v_CA21_4317:** Median monthly shelter costs for rented dwellings (\$)\

**Hover, click CT for more info, zoom in/out, and toggle between the two choropleth layers (top right)**
```{r, echo=FALSE}
map_leaflet_med_cost
```

### Table
**v_CA21_4309:** Median monthly shelter costs for owned dwellings (\$)\
**v_CA21_4310:** Average monthly shelter costs for owned dwellings (\$)\
**v_CA21_4317:** Median monthly shelter costs for rented dwellings (\$)\
**v_CA21_4318:** Average monthly shelter costs for rented dwellings (\$)\

**To sort by a variable, click on the variable name**
```{r, warning=FALSE,echo=FALSE}
table_costs
```

# Appendix
The following tables and figures support the visualizations above.

## Census Variable Lookup{#var-lookup}
This table allows you to search through all 2021 Census variables by name,
keyword or by 'vector', which is a naming convention from the [cancensus CensusMapper API](https://mountainmath.github.io/cancensus/articles/data_discovery.html) not 
from StatsCan. 

Search by vector code for full details of any variable shown below, or use this
interactive table to explore the full range of census variables.

**Try the search bar**
```{r, warning=FALSE,echo=FALSE}
table_var_lookup
```

## Naming Census Tracts{#naming-cts}

When I began looking at Census data on StatsCan's sites and then using Cancensus,
a major pain point was that the Census Tracts are named by numbers. It is common
for Census data to be aggregated and presented at a smaller scale such as 
by neighbourhood, smaller scale meaning larger geographic units of analysis (which is confusing I know!),
but in the context of Toronto's heterogeneity, I wanted to keep 
the fine grained detail of the CTs but provide CTs with meaningful names.

To do this, I calculated the centroid (blue dots below) of each CT polygon, then
determined in which neighbourhood each CT fell based on these centroids. 
The neighbourhood name of each CT was then concatenated to the CT number name. 
The CTs with their new names are shown on this interactive map, as are the layers 
used to name them. The logic for using centroids came from the [Ontario Community Health Profiles Partnership](https://www.ontariohealthprofiles.ca/resourcesON.php#convfiles), thank you!

```{r, out.width = "100%", echo=FALSE}

# To check, mobile responsiveness is improved now, but to test if out.width
# here is helping or if it was the width var in map_leaflet_perc30 only

map_naming_cts
```


## Process Notes
This analysis was done entirely in R, leaning heavily on the Cancensus, Leaflet, 
Tidyverse, and Plotly packages among others. Many thanks to all the talented 
open source developers who build and maintain these amazing tools!

Note that it would be easy to build a Shiny or Dash web application that allows 
the user to interactivley update any of these visualizations with any variable,
however, the benefit of this style of document is that it opens on any personal
computer without getting a server involved, or can be hosted online. The downside 
to this format however is a very large file size compared to a reactive solution using a server. 

This report was developed for desktop viewing, but it is responsive and looks
good on your smaller screens too.

If you have accessibility concerns, please reach out and I'll make whatever improvements I can.

# Contact

If you have any questions, comments, or feedback, I'm at fraser.mclaughlin@outlook.com.
I specialize in data analysis, visualization and building decision making tools. Reach out 
if you'd like to collaborate!

I hope you've found this useful.\
&copy; Fraser McLaughlin 2023\